What is happening on Twitter? Tweet analysis and student research projects

### 2020-05-31 Rmd/jse-special-issue-2020.Rmd
added paragraph on inference for LDA

Twitter has profoundly changed how we communicate. 
In only 280 characters, users instantly contribute to public conversations on politics, 
current events, sports, media, and many other topics. 
Recent development of accessible statistical methods for large-scale text 
analysis now enable instructors to use tweets as contemporary pedagogical 
tools in guiding undergraduate research projects.
We report one instance of a mentored text analysis research project. 
We share our data and computer code to encourage others to undertake 
tweet text analysis research. We also describe our methods for creating a 
collection of tweets.

@@

While our tweet collection enables us to address many possible research questions, the 
dynamic content of tweets over time particularly piqued our interest. Together, students and 
mentors hypothesized that high-profile social media events would generate a high volume of 
tweets, and that we would detect social media events through changes in tweet 
topic content over time. We present below 1) an approach for collecting tweets in real 
time and 2) methods for detecting social media 
events via latent Dirichlet allocation modeling of tweets and 3) suggestions for 
using tweets in research mentoring of undergraduate students.

@@

1. Write R code to perform text analysis of large volumes of tweets.
1. Communicate results in a written report and poster presentation.
1. Translate statistical findings into scientific conclusions.
1. Develop data science strategies to address a scientific research question.

@@

We collaboratively developed research goals with students through a series of discussions throughout the academic year. As trainees begin their senior research projects, we suggest that mentors discuss with them:

@@

```{bash, eval = FALSE}
*/5 * * * * R -e 'rtweet::stream_tweets(timeout = (60 * 5), 
parse = FALSE, file_name = paste0("~/work/mentoring/mentoring-framework/data/",
lubridate::now(), "-tweets"))'
```

@@

Below is an example of Tweet JSON. Every tweet features the keys "created\_at" (the time stamp), 
"id\_str" (a unique tweet identifier), and "text". We use these three keys in 
our analyses.

@@

Latent Dirichlet allocation models require that the corpus be inputted as a 
document by term matrix. In a document by term matrix, each row corresponds to a single document (a single 
tweet), and each column is a single term (or word). Each cell contains a 
count (the number of 
occurrences of a term in the specified document). We created a document by 
term matrix with the R function `cast_dtm` from the `tidytext` package.

@@

Latent Dirichlet allocation is a statistical method for inferring latent (unobservable) topics (or themes)
from a large corpus (or collection) of documents [@blei2003latent].
We pretend that there's an imaginary process for creating documents in the corpus. For each document,
we choose a discrete distribution over topics. For example, some tweets from Memorial Day may 
refer to the holiday. This may constitute one topic in the corpus. 
Having chosen a distribution over topics, we then select document words by first 
drawing a topic from the distribution over topics, then drawing a word from the
chosen topic. 
Thus, a topic is technically defined as a distribution over words in a fixed vocabulary (or collection of words) [@blei2003latent].

@@

Tweet analysis gives students practical experience in the data science process of 
formulating a research question, gathering data to address it, summarizing the 
data, visualizing results, and communicating findings.

>>>>

Twitter has profoundly changed how we communicate. 
In only 280 characters, users instantly contribute to public conversations on politics, 
current events, sports, media, and many other topics. 
Recent development of accessible statistical methods for large-scale text 
analysis now enable instructors to use tweets as contemporary pedagogical 
tools in guiding undergraduate research projects.
We report mentored text analysis projects. 
We share our data and computer code to encourage others to undertake 
tweet text analysis research. We also describe our methods for creating a 
collection of tweets.

@@

While our tweet collection enables us to address many possible research questions, the 
dynamic content of tweets over time particularly piqued our interest. Together, students and 
mentors hypothesized that high-profile social media events would generate a high volume of 
tweets, and that we would detect social media events through changes in tweet 
topic content over time. We discuss in detail below one approach to studying this
question.

In the sections that follow, we detail our backward design-inspired approach to 
writing learning objectives, preliminary research mentoring considerations, 
data science methods for collecting and analyzing tweets, analysis results, 
and thoughts on assessment and advanced topics.

@@

1. Write R code to perform text analysis of large volumes of tweets [@r].
1. Communicate results in a written report and poster presentation.
1. Translate statistical findings into scientific conclusions.
1. Develop data science strategies to address a scientific research question.

@@

We collaboratively developed research goals with students in a series of discussions during the academic year. As trainees begin their senior research projects, we suggest that mentors discuss with them:

@@

```{bash, eval = FALSE, echo = TRUE}
*/5 * * * * R -e 'rtweet::stream_tweets(timeout = (60 * 5), 
parse = FALSE, file_name = paste0("~/work/mentoring/mentoring-framework/data/",
lubridate::now(), "-tweets"))'
```

@@

Below is an example of Tweet JSON.

@@

Latent Dirichlet allocation models require that the corpus be organized as a 
document by term matrix. In a document by term matrix, each row corresponds to a single document (a single 
tweet), and each column is a single term (or word). Each cell contains a 
count (the number of 
occurrences of a term in the specified document). We created a document by 
term matrix with the R function `cast_dtm` from the `tidytext` package.

@@

Latent Dirichlet allocation is a statistical method for inferring latent (unobservable) topics (or themes)
from a large corpus (or collection) of documents [@blei2003latent].
We pretend that there's an imaginary process for creating documents in the corpus. For each document,
we choose a discrete distribution over topics. For example, some tweets from Memorial Day may 
refer to the holiday. This may constitute one topic in the corpus. 
Having chosen a distribution over topics, we then select document words by first 
drawing a topic from the distribution over topics, then drawing a word from the
chosen topic. 
The goal for latent Dirichlet allocation is to infer both the distribution over topics and the topics [@blei2003latent]. A topic, in this setting, is a distribution over 
the vocabulary (the collection of all words in a corpus).

Inference for latent Dirichlet allocation model is performed by either 
sampling from the posterior distribution or through variational methods. Researchers
have devised a variety of Gibbs sampling techniques for latent Dirichlet allocation models. 
Variational methods, while using approximations to the posterior distribution, offer the 
advantage of computational speed. We used variational methods below.

@@

Tweet analysis gives students practical experience in the data science process of 
formulating a research question, gathering data to address it, summarizing the 
data, visualizing results, and communicating findings. By considering first student 
research interests and integrating them with our senior thesis learning objectives, 
we successfully guided two undergraduate researchers in data science research with
tweets.
